Reverse eigenvalue problem

### 2024-03-19 Docs/NhanesApplExample.Rmd
add intercept

# Spline basis

@@

```{r set_up}
# basis to try
bs_vec <- c("cr", "cs", "tp", "bs", "ps")
evals <- matrix(NA, length(bs_vec), K)
```

@@

```{r try_splines, results='hide'}
pb <- txtProgressBar(min=0, max=length(bs_vec), style = 3)

t1 <- Sys.time()
for(i in seq_along(bs_vec)){

this_glmm <- bam(Y ~ s(sind, bs = bs_vec[i])+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"),
                   family = binomial,
                   data=train_df,
                   method = "fREML",
                   discrete = TRUE)
evals[i, ] <- 1/this_glmm$sp[2:5]/n_bin
setTxtProgressBar(pb, i)

}

close(pb)
t2 <- Sys.time()
# t2-t1 # about 5 minutes
```

```{r}
data.frame(bs_vec, evals)
```

@@

# mean 
true_mu <- predict(debias_glmm, type = "terms")[, 1]
true_mu <- matrix(true_mu, nrow = J)[, 1]
# plot(1:J, true_mu)

@@

This is the first discrepancy I have found - It looks like the simulated data is more active and more variable than the real data. Doesn't that mean the algorithm overestimates the population activity and variability? Since the mean estimates will surely change in Step 4 (debias), would the overestimation be caused by that?

@@

# Step 2:  Local GLMM
gen_bin_lst <- split(gen_df, f = gen_df$bin)
df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)
## Figure
# df_est_latent %>% filter(id %in% 1:4) %>%
#   ggplot()+
#   geom_point(aes(x=sind, y=Y), size = 0.5)+
#   geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
#   geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
#   facet_wrap(~id)

@@

## debias
est_phi <- data.frame(sind = 1:J, est_phi)
colnames(est_phi) <- c("sind", paste0("phi", 1:K))
gen_df <- gen_df %>% left_join(est_phi, by = "sind")
gen_df$id <- as.factor(gen_df$id)
### model fit with B-splines basis
t1 <- Sys.time()
this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
t2 <- Sys.time()
```

@@

```{r compare_mean}
bind_rows(
  data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"),
  data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]) %>% mutate(type = "BAM"),
  data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA")
) %>%
  ggplot()+
      geom_line(aes(x=sind, y=mu, col = type))+
      labs(title="Mean estimates")

# Simulation with the FPCA mean

I would like to experiments if I could simulated data whose activity level and variation is closer to NHANES, using the interpolated FPCA mean and un-evaluated eigenvalues.

```{r load_fpca_mu}
# load FPCA model from NHANES
load(here("Data/Appl_fpca_model.RData"))

# compare means on NHANES data from FPCA and debias model
bind_rows(
  data.frame(sind=1:J, mu = true_mu, type = "BAM"),
  data.frame(sind=mid, mu = fpca_mod$mu, type = "FPCA")
) %>% 
  ggplot()+
  geom_line(aes(x=sind, y=mu, col = type))+
  labs(title = "Mean estimates on the NHANES dataset")

```{r gen_data_fpca_mu}
# interpolate FPCA mean
interp_mu_fpca <- approx(mid, fpca_mod$mu, xout=1:J, rule=2)
true_mu <- interp_mu_fpca$y

# FPCA eigenvalues
true_evals <- fpca_mod$evalues/n_bin

# maybe increase sample size?
# Ntry <- 1000

# generate score
xi <- rmvnorm(Ntry, mean = rep(0, K), sigma = diag(true_evals))

# add in un-evaluated mean
eta <- apply(bi, 2, function(x){x+true_mu})
gen_df <- data.frame(id = factor(rep(1:Ntry, each=J)),
                     sind = rep(1:J, Ntry),
                     eta_i = as.vector(eta))

```{r plot_data_fpca_mu}
# comparing generated data to the actual data
bind_rows(
    gen_df %>% 
      group_by(sind) %>% 
      summarize(active = mean(Y)) %>%
      mutate(type="Simulation"),

Looks like by using the FPCA mean, I was able to lower the activity level of generated data down to the level of NHANES data.

```{r fgfpca_fpca_mean}
# bin data
# using the same bin length 
gen_df$bin <- cut(gen_df$sind, breaks = brks, include.lowest = T, labels = mid)
gen_df$bin <- as.numeric(as.character(gen_df$bin))

# Step 2:  Local GLMM
gen_bin_lst <- split(gen_df, f = gen_df$bin)
df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)
## Figure
# df_est_latent %>% filter(id %in% 1:4) %>%
#   ggplot()+
#   geom_point(aes(x=sind, y=Y), size = 0.5)+
#   geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
#   geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
#   facet_wrap(~id)

# FPCA
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                        nrow=Ntry, 
                        ncol=n_bin, byrow = F) 
# dim(mat_est_unique)
fpca_mod_sim <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K) # keep 4 PCs
```

B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
                outer.ok = TRUE)$design  # evaluate B-splines on binned grid

Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
                   outer.ok = TRUE)$design
for(k in 1:K){
  lm_mod <- lm(fpca_mod_sim$efunctions[,k] ~ B-1)
  est_phi[,k] <- Bnew %*% coef(lm_mod)
}# project binned eigenfunctions onto the original grid

## debias
est_phi <- data.frame(sind = 1:J, est_phi)
colnames(est_phi) <- c("sind", paste0("phi", 1:K))
gen_df <- gen_df %>% left_join(est_phi, by = "sind")
gen_df$id <- as.factor(gen_df$id)
### model fit with B-splines basis
t1 <- Sys.time()
this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
t2 <- Sys.time()
```

Estimated eigenvalues

```{r}
1/this_glmm$sp[2:5]/sqrt(n_bin)
```

Using the FPCA estimates, I was able to lower the activity but not variation. I was also not able to recreate the reversed eigenvalue problem. Also, singularity issue happened in the binnin step when using the FPCA estimates to generate data.

>>>>

# Truncated track

@@

@@

```{r try_splines, results='hide'}
# pb <- txtProgressBar(min=0, max=length(bs_vec), style = 3)
# 
# t1 <- Sys.time()
# for(i in seq_along(bs_vec)){

this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"),
                   family = binomial,
                   data=train_df,
                   method = "fREML",
                   discrete = TRUE)
# evals[i, ] <- 1/this_glmm$sp[2:5]/n_bin
# setTxtProgressBar(pb, i)
# 
# }
# 
# close(pb)
# t2 <- Sys.time()
# t2-t1 # about 5 minutes
```

```{r, class.source='fold-show'}
# re-evaluated eigenvalues
1/this_glmm$sp[2:5]/n_bin
```

@@

# mean 
# because of the parametrization of mgcv
# we need to add the constant intercept back
true_mu <- predict(debias_glmm, type = "terms")[, 1]+coef(debias_glmm)[1]
true_mu <- matrix(true_mu, nrow = J)[, 1]
# plot(1:J, true_mu)

@@

After adding the intercept back, the generated data is still slightly more active and variable than the true data. Is it enough to make a difference?

@@

# Step 2:  Local GLMM
gen_bin_lst <- split(gen_df, f = gen_df$bin)
df_est_latent <- lapply(gen_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
df_est_latent <- bind_rows(df_est_latent)
## Figure
df_est_latent %>% filter(id %in% 1:4) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.5)+
  geom_line(aes(x=sind, y=plogis(eta_i), col = "true"))+
  geom_line(aes(x=sind, y=plogis(eta_hat), col = "estimate"))+
  facet_wrap(~id)

@@

## debias
est_phi <- data.frame(sind = 1:J, est_phi)
colnames(est_phi) <- c("sind", paste0("phi", 1:K))
gen_df <- gen_df %>% left_join(est_phi, by = "sind")
gen_df$id <- as.factor(gen_df$id)

### model fit with B-splines basis
t1 <- Sys.time()
this_glmm <- bam(Y ~ s(sind, bs = "cr")+
                   s(id, by=phi1, bs="re")+
                   s(id, by=phi2, bs="re")+
                   s(id, by=phi3, bs="re")+
                   s(id, by=phi4, bs="re"),
                 family = binomial,
                 data=gen_df,
                 method = "fREML",
                 discrete = TRUE)
t2 <- Sys.time()
```

@@

```{r compare_mean}
bind_rows(
  data.frame(sind=1:J, mu=true_mu) %>% mutate(type = "True"),
  data.frame(sind=1:J, mu=predict(this_glmm, type = "terms")[1:J, 1]+coef(this_glmm)[1]) %>% mutate(type = "BAM"),
  data.frame(sind=mid, mu=fpca_mod$mu) %>% mutate(type="FPCA")
) %>%
  ggplot()+
      geom_line(aes(x=sind, y=mu, col = type))+
      labs(title="Mean estimates")
```
